---
title: "Analysis of Multilingual Models Performance"
author: "Sean Trott and Pam Rivière"
date: "November 20, 2024"
output:
  html_document:
    keep_md: yes
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 300, fig.format = "png")
```


```{r include=FALSE}
library(tidyverse)
library(lme4)
library(viridis)
library(ggridges)
library(broom)
library(lmerTest)
library(ggrepel)

all_colors <- viridis::viridis(10, option = "magma")
my_colors <- all_colors[c(1, 4, 7, 9)]  # Selecting specific colors from the palette
```

# Original models

## Load data


```{r}
# setwd("/Users/seantrott/Dropbox/UCSD/Research/Ambiguity/SSD/multilingual_disambiguation/src")
directory_path <- "../data/processed/isotropy/"
csv_files <- list.files(path = directory_path, pattern = "*.csv", full.names = TRUE)
csv_list <- csv_files %>%
  map(~ read_csv(.))
df_original_models <- bind_rows(csv_list) %>%
  filter(Layer > 0) %>%
  mutate(
    model_name = sub(".*/", "", mpath)
  ) %>%
  mutate(language = str_to_title(language))

table(df_original_models$language)
table(df_original_models$multilingual)
table(df_original_models$model_name)


```

## Visualize isotropy metrics



```{r}

### 
df_layerwise_isotropy = df_original_models %>%
  group_by(model_name, n_params, Layer, language, multilingual) %>%
  summarise(mean_dist = mean(mean_dist),
            mean_centered_isotropy = mean(centered_isotropy),
            mean_intra_sentence_similarity = mean(intra_sentence_similarity)) 

# Summarize multilingual vs. not
summary_df <- df_layerwise_isotropy %>%
  group_by(multilingual, Layer, language) %>%
  summarise(mean_dist = mean(mean_dist),
            mean_centered_isotropy = mean(mean_centered_isotropy),
            mean_intra_sentence_similarity = mean(mean_intra_sentence_similarity))



df_layerwise_isotropy %>%
  ggplot(aes(x = Layer, y = mean_dist, color = model_name)) +
  geom_line(alpha = 0.6, linetype = "dotted") +
  scale_color_manual(values = viridisLite::viridis(10, option = "mako", 
                                                   begin = 0.8, end = 0.15)) + 
  geom_line(data = summary_df,
            aes(x = Layer, 
                y = mean_dist, 
                group = multilingual, 
                color = multilingual),
            inherit.aes = FALSE,
            size = 2) +
  scale_color_manual(
    values = c("Yes" = viridisLite::viridis(2, option = "mako", begin = 0.8, end = 0.15)[1],
               "No"  = viridisLite::viridis(2, option = "mako", begin = 0.8, end = 0.15)[2])
  ) +
  theme_minimal() +
  labs(x = "Layer", y = "Mean Cosine Distance", color = "Multilingual") +
  theme(text = element_text(size = 15),
        legend.title = element_blank()) +
  facet_wrap(~language)


df_layerwise_isotropy %>%
  ggplot(aes(x = Layer, y = mean_centered_isotropy, color = model_name)) +
  geom_line(alpha = 0.6, linetype = "dotted") +
  scale_color_manual(values = viridisLite::viridis(10, option = "mako", 
                                                   begin = 0.8, end = 0.15)) + 
  geom_line(data = summary_df,
            aes(x = Layer, 
                y = mean_centered_isotropy, 
                group = multilingual, 
                color = multilingual),
            inherit.aes = FALSE,
            size = 2) +
  scale_color_manual(
    values = c("Yes" = viridisLite::viridis(2, option = "mako", begin = 0.8, end = 0.15)[1],
               "No"  = viridisLite::viridis(2, option = "mako", begin = 0.8, end = 0.15)[2])
  ) +
  theme_minimal() +
  labs(x = "Layer", y = "Mean Centered Isotropy", color = "Multilingual") +
  theme(text = element_text(size = 15),
        legend.title = element_blank()) +
  facet_wrap(~language)


df_layerwise_isotropy %>%
  ggplot(aes(x = Layer, y = mean_intra_sentence_similarity, color = model_name)) +
  geom_line(alpha = 0.6, linetype = "dotted") +
  scale_color_manual(values = viridisLite::viridis(10, option = "mako", 
                                                   begin = 0.8, end = 0.15)) + 
  geom_line(data = summary_df,
            aes(x = Layer, 
                y = mean_intra_sentence_similarity, 
                group = multilingual, 
                color = multilingual),
            inherit.aes = FALSE,
            size = 2) +
  scale_color_manual(
    values = c("Yes" = viridisLite::viridis(2, option = "mako", begin = 0.8, end = 0.15)[1],
               "No"  = viridisLite::viridis(2, option = "mako", begin = 0.8, end = 0.15)[2])
  ) +
  theme_minimal() +
  labs(x = "Layer", y = "Mean Intra-Sentence Similarity", color = "Multilingual") +
  theme(text = element_text(size = 15),
        legend.title = element_blank()) +
  facet_wrap(~language)

### Max by size
df_models_isotropy_by_params = df_original_models %>%
  group_by(model_name, multilingual, language, n_params, Layer) %>%
  summarise(mean_dist = mean(mean_dist),
            mean_centered_isotropy = mean(centered_isotropy),
            mean_intra_sentence_similarity = mean(intra_sentence_similarity)) %>%
  group_by(model_name, n_params, multilingual, language) %>%
  summarise(max_dist = max(mean_dist),
            max_centered_isotropy = max(mean_centered_isotropy),
            min_iss = min(mean_intra_sentence_similarity)) %>%
  mutate(log_params = log10(n_params))

df_models_isotropy_by_params %>%
  ggplot(aes(x = n_params,
             y = max_dist,
             color = multilingual,
             shape = multilingual)) +
  geom_point(size = 6,
             alpha = .5) +
  scale_x_log10() +
  geom_text_repel(aes(label=model_name), size=3) +
  labs(x = "Parameters",
       y = "Max Mean Cosine Distance",
       color = "Multilingual",
       shape = "Multilingual") +
  theme_bw() +
  scale_color_manual(
    values = c("Yes" = viridisLite::viridis(2, option = "mako", begin = 0.8, end = 0.15)[1],
               "No"  = viridisLite::viridis(2, option = "mako", begin = 0.8, end = 0.15)[2])
  ) +
  theme(text = element_text(size = 15),
        legend.position="bottom") +
  facet_wrap(~language)

df_models_isotropy_by_params %>%
  ggplot(aes(x = n_params,
             y = max_centered_isotropy,
             color = multilingual,
             shape = multilingual)) +
  geom_point(size = 6,
             alpha = .5) +
  scale_x_log10() +
  geom_text_repel(aes(label=model_name), size=3) +
  labs(x = "Parameters",
       y = "Max Centered Isotropy",
       color = "Multilingual",
       shape = "Multilingual") +
  theme_bw() +
  scale_color_manual(
    values = c("Yes" = viridisLite::viridis(2, option = "mako", begin = 0.8, end = 0.15)[1],
               "No"  = viridisLite::viridis(2, option = "mako", begin = 0.8, end = 0.15)[2])
  ) +
  theme(text = element_text(size = 15),
        legend.position="bottom") +
  facet_wrap(~language)


df_models_isotropy_by_params %>%
  ggplot(aes(x = n_params,
             y = min_iss,
             color = multilingual,
             shape = multilingual)) +
  geom_point(size = 6,
             alpha = .5) +
  scale_x_log10() +
  geom_text_repel(aes(label=model_name), size=3) +
  labs(x = "Parameters",
       y = "Min Intra-Sentence Similarity",
       color = "Multilingual",
       shape = "Multilingual") +
  theme_bw() +
  scale_color_manual(
    values = c("Yes" = viridisLite::viridis(2, option = "mako", begin = 0.8, end = 0.15)[1],
               "No"  = viridisLite::viridis(2, option = "mako", begin = 0.8, end = 0.15)[2])
  ) +
  theme(text = element_text(size = 15),
        legend.position="bottom") +
  facet_wrap(~language)


### Mean by size
df_models_isotropy_by_params = df_original_models %>%
  group_by(model_name, multilingual, language, n_params) %>%
  summarise(mean_dist = mean(mean_dist),
            mean_centered_isotropy = mean(centered_isotropy),
            mean_iss = mean(intra_sentence_similarity)) %>%
  mutate(log_params = log10(n_params))

df_models_isotropy_by_params %>%
  ggplot(aes(x = n_params,
             y = mean_dist,
             color = multilingual,
             shape = multilingual)) +
  geom_point(size = 6,
             alpha = .5) +
  scale_x_log10() +
  geom_text_repel(aes(label=model_name), size=3) +
  labs(x = "Parameters",
       y = "Mean Cosine Distance",
       color = "Multilingual",
       shape = "Multilingual") +
  theme_bw() +
  scale_color_manual(
    values = c("Yes" = viridisLite::viridis(2, option = "mako", begin = 0.8, end = 0.15)[1],
               "No"  = viridisLite::viridis(2, option = "mako", begin = 0.8, end = 0.15)[2])
  ) +
  theme(text = element_text(size = 15),
        legend.position="bottom") +
  facet_wrap(~language)

df_models_isotropy_by_params %>%
  ggplot(aes(x = n_params,
             y = mean_centered_isotropy,
             color = multilingual,
             shape = multilingual)) +
  geom_point(size = 6,
             alpha = .5) +
  scale_x_log10() +
  geom_text_repel(aes(label=model_name), size=3) +
  labs(x = "Parameters",
       y = "Mean Centered Isotropy",
       color = "Multilingual",
       shape = "Multilingual") +
  theme_bw() +
  scale_color_manual(
    values = c("Yes" = viridisLite::viridis(2, option = "mako", begin = 0.8, end = 0.15)[1],
               "No"  = viridisLite::viridis(2, option = "mako", begin = 0.8, end = 0.15)[2])
  ) +
  theme(text = element_text(size = 15),
        legend.position="bottom") +
  facet_wrap(~language)

df_models_isotropy_by_params %>%
  ggplot(aes(x = n_params,
             y = mean_iss,
             color = multilingual,
             shape = multilingual)) +
  geom_point(size = 6,
             alpha = .5) +
  scale_x_log10() +
  geom_text_repel(aes(label=model_name), size=3) +
  labs(x = "Parameters",
       y = "Mean Intra-Sentence Similarity",
       color = "Multilingual",
       shape = "Multilingual") +
  theme_bw() +
  scale_color_manual(
    values = c("Yes" = viridisLite::viridis(2, option = "mako", begin = 0.8, end = 0.15)[1],
               "No"  = viridisLite::viridis(2, option = "mako", begin = 0.8, end = 0.15)[2])
  ) +
  theme(text = element_text(size = 15),
        legend.position="bottom") +
  facet_wrap(~language)

##### Layer depth ratio
df_layerwise_isotropy = df_layerwise_isotropy %>%
  group_by(model_name) %>%
  mutate(max_layer = max(Layer),
         prop_layer = Layer / max_layer)


df_layerwise_isotropy %>%
  mutate(binned_prop_layer = ntile(prop_layer, 10)) %>%
  mutate(prop_binned = binned_prop_layer / 10) %>%
  ggplot(aes(x = prop_binned,
             y = mean_centered_isotropy)) +
  stat_summary(
    aes(group = multilingual,
        color = multilingual),  
    fun = mean,    
    geom = "line",        
    size = 2              
  ) +
  stat_summary(
    aes(group = multilingual, 
        fill = multilingual), 
    fun.data = mean_se,    
    geom = "ribbon",  
    alpha = 0.2,   
    color = NA     
  ) +
  theme_minimal() +
  labs(x = "Layer Depth Ratio",
       y = "Mean Centered Isotropy",
  color = "Multilingual",
  fill = "Multilingual") +
  scale_color_manual(
    values = c("Yes" = viridisLite::viridis(2, option = "mako", begin = 0.8, end = 0.15)[1],
               "No"  = viridisLite::viridis(2, option = "mako", begin = 0.8, end = 0.15)[2])
  ) +
  theme(text = element_text(size = 15),
        legend.position = "bottom")  +
  facet_wrap(~language)



df_layerwise_isotropy %>%
  mutate(binned_prop_layer = ntile(prop_layer, 10)) %>%
  mutate(prop_binned = binned_prop_layer / 10) %>%
  ggplot(aes(x = prop_binned,
             y = mean_dist)) +
  stat_summary(
    aes(group = multilingual,
        color = multilingual),  
    fun = mean,    
    geom = "line",        
    size = 2              
  ) +
  stat_summary(
    aes(group = multilingual, 
        fill = multilingual), 
    fun.data = mean_se,    
    geom = "ribbon",  
    alpha = 0.2,   
    color = NA     
  ) +
  theme_minimal() +
  labs(x = "Layer Depth Ratio",
       y = "Mean Cosine Distance",
  color = "Multilingual",
  fill = "Multilingual") +
  scale_color_manual(
    values = c("Yes" = viridisLite::viridis(2, option = "mako", begin = 0.8, end = 0.15)[1],
               "No"  = viridisLite::viridis(2, option = "mako", begin = 0.8, end = 0.15)[2])
  ) +
  theme(text = element_text(size = 15),
        legend.position = "bottom")  +
  facet_wrap(~language)


df_layerwise_isotropy %>%
  mutate(binned_prop_layer = ntile(prop_layer, 10)) %>%
  mutate(prop_binned = binned_prop_layer / 10) %>%
  ggplot(aes(x = prop_binned,
             y = mean_intra_sentence_similarity)) +
  stat_summary(
    aes(group = multilingual,
        color = multilingual),  
    fun = mean,    
    geom = "line",        
    size = 2              
  ) +
  stat_summary(
    aes(group = multilingual, 
        fill = multilingual), 
    fun.data = mean_se,    
    geom = "ribbon",  
    alpha = 0.2,   
    color = NA     
  ) +
  theme_minimal() +
  labs(x = "Layer Depth Ratio",
       y = "Mean Intra-Sentence Similarity",
  color = "Multilingual",
  fill = "Multilingual") +
  scale_color_manual(
    values = c("Yes" = viridisLite::viridis(2, option = "mako", begin = 0.8, end = 0.15)[1],
               "No"  = viridisLite::viridis(2, option = "mako", begin = 0.8, end = 0.15)[2])
  ) +
  theme(text = element_text(size = 15),
        legend.position = "bottom")  +
  facet_wrap(~language)


### Isotropy by layer depth ratio
df_original_models = df_original_models %>%
  group_by(model_name) %>%
  mutate(max_layer = max(Layer),
         prop_layer = Layer / max_layer) %>%
  ### Scale for interpreting the coefficients more easily
  mutate(prop_layer_scaled = scale(prop_layer))

mod_full = lmer(data = df_original_models,
                mean_dist ~ prop_layer_scaled * multilingual + 
                  language + 
                  log10(n_params) +
                  (1 | word) + (1|model_name),
                REML = FALSE)
summary(mod_full)

mod_full = lmer(data = df_original_models,
                centered_isotropy ~ prop_layer_scaled * multilingual + 
                  language + 
                  log10(n_params) +
                  (1 | word) + (1 |model_name),
                REML = FALSE)
summary(mod_full)

mod_full = lmer(data = df_original_models,
                intra_sentence_similarity ~ prop_layer_scaled * multilingual + 
                  language + 
                  log10(n_params) +
                  (1 | word) + (1 |model_name),
                REML = FALSE)
summary(mod_full)

```


## Does isotropy predict R2?


**NOTES**:

- Possible collinearity between predictors. Intra-sentence similarity is way too similar to mean distance.
- Different predictors go in diff. directions / implications?
- Trying out mediation analysis...
- Taht said, we see the opposite effect with `mean_dist`, b/c higher `mean_Dist` is correlated with lower R2...

```{r}
### Load data and calculate $R^2$
directory_path <- "../data/processed/distances/"
csv_files <- list.files(path = directory_path, pattern = "*.csv", full.names = TRUE)
csv_list <- csv_files %>%
  map(~ read_csv(.))
df_distances <- bind_rows(csv_list) %>%
  mutate(
    model_name = sub(".*/", "", mpath)
  ) %>%
  mutate(language = str_to_title(language))


df_original_models_r2 = df_distances %>%
  group_by(model_name, multilingual, language, n_params, Layer) %>%
  summarise(r = cor(Distance, mean_relatedness)) %>%
  mutate(r2 = r ** 2) 



#### Merge with isotropy
df_layerwise_isotropy_r2 = df_layerwise_isotropy %>%
  inner_join(df_original_models_r2)

### Mean centered isotropy
mod_full = lmer(data = df_layerwise_isotropy_r2,
                r2 ~ mean_centered_isotropy +
                  prop_layer +
                  multilingual + 
                  language + 
                  log10(n_params) +
                   (1 |model_name),
                REML = FALSE)
summary(mod_full)

### Mean cosine distance
mod_full = lmer(data = df_layerwise_isotropy_r2,
                r2 ~ mean_dist +
                  prop_layer +
                  multilingual + 
                  language + 
                  log10(n_params) +
                   (1 |model_name),
                REML = FALSE)
summary(mod_full)

### intra-sentence similarity
mod_full = lmer(data = df_layerwise_isotropy_r2,
                r2 ~ mean_intra_sentence_similarity +
                  prop_layer +
                  multilingual + 
                  language + 
                  log10(n_params) +
                   (1 |model_name),
                REML = FALSE)
summary(mod_full)



#### Plotting z-scores
df_layerwise_isotropy_r2 = df_layerwise_isotropy_r2 %>%
  group_by(model_name) %>%
  mutate(mean_iss_z = scale(mean_intra_sentence_similarity),
         mean_r2_z = scale(r2),
         mean_ci_z = scale(mean_centered_isotropy),
         mean_dist_z = scale(mean_dist)) 

df_long <- df_layerwise_isotropy_r2 %>%
  pivot_longer(cols = c(mean_iss_z, mean_r2_z),
               names_to = "metric",
               values_to = "value")

df_long %>%
  ggplot(aes(x = Layer, y = value, color = metric)) +
  geom_line() +
  geom_point(alpha = 0.6) +
  facet_wrap(~model_name) +
  theme_minimal() +
  labs(y = "Value", x = "Layer", color = "Metric")

df_long <- df_layerwise_isotropy_r2 %>%
  pivot_longer(cols = c(mean_ci_z, mean_r2_z),
               names_to = "metric",
               values_to = "value")

df_long %>%
  ggplot(aes(x = Layer, y = value, color = metric)) +
  geom_line() +
  geom_point(alpha = 0.6) +
  facet_wrap(~model_name) +
  theme_minimal() +
  labs(y = "Value", x = "Layer", color = "Metric")

df_long <- df_layerwise_isotropy_r2 %>%
  pivot_longer(cols = c(mean_dist_z, mean_r2_z),
               names_to = "metric",
               values_to = "value")
df_long %>%
  ggplot(aes(x = Layer, y = value, color = metric)) +
  geom_line() +
  geom_point(alpha = 0.6) +
  facet_wrap(~model_name) +
  theme_minimal() +
  labs(y = "Value", x = "Layer", color = "Metric")


df_long <- df_layerwise_isotropy_r2 %>%
  pivot_longer(cols = c(mean_iss_z, mean_ci_z),
               names_to = "metric",
               values_to = "value")
df_long %>%
  ggplot(aes(x = Layer, y = value, color = metric)) +
  geom_line() +
  geom_point(alpha = 0.6) +
  facet_wrap(~model_name) +
  theme_minimal() +
  labs(y = "Value", x = "Layer", color = "Metric")


#### Mediation analysis

mod_A <- lmer(r2 ~ multilingual + 
                language + log10(n_params) + 
                (1 | model_name), 
              data = df_layerwise_isotropy_r2)

mod_B <- lmer(mean_centered_isotropy ~ multilingual + 
                (1 | model_name), data = df_layerwise_isotropy_r2)

mod_C <- lmer(r2 ~ multilingual + mean_centered_isotropy + 
                  prop_layer +
                  language + 
                  log10(n_params) +
                (1 | model_name), 
              data = df_layerwise_isotropy_r2)

coef_A <- fixef(mod_A)["multilingualYes"]
coef_C <- fixef(mod_C)["multilingualYes"]

# Proportion of multilingual effect explained by adding anisotropy
prop_explained <- (coef_A - coef_C) / coef_A
prop_explained


### Logic: 
### m_boot captures effect of multilingual on CI
### y_boot captures effect of CI on R2, independent of multilingual status
### then we capture the effect of multilingual on R2 *through* CI by multiplying these together
set.seed(1)
n_boot <- 100
boot_indirect <- replicate(n_boot, {
  sample_rows <- sample(nrow(df_layerwise_isotropy_r2), replace = TRUE)
  df_boot <- df_layerwise_isotropy_r2[sample_rows, ]

  mod_m_boot <- lmer(mean_centered_isotropy ~ multilingual + (1 | model_name), data = df_boot)
  mod_y_boot <- lmer(r2 ~ multilingual + mean_centered_isotropy + 
                  prop_layer +
                  language + 
                  log10(n_params) +
                (1 | model_name), 
              data = df_boot)
  
  a <- fixef(mod_m_boot)["multilingualYes"]
  b <- fixef(mod_y_boot)["mean_centered_isotropy"]
  
  a * b
})

# Confidence interval
quantile(boot_indirect, probs = c(0.025, 0.975))

# Optional: mean indirect effect: How much does being multilingual lower R2 by means of mean centered isotropy
mean(boot_indirect)
```

